Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2021-11-17 paper/latex/generic/notes.tex
get generic article format working with agsm style

\documentclass[NETN,manuscript]{stjour-new}
\articletype{Research}
\usepackage{graphicx}
\usepackage{cleveref}
\graphicspath{ {figs/} }
% \input{preamble.tex}
% \documentclass{article}
% \usepackage[utf8]{inputenc}
% \usepackage{geometry}
% \usepackage{amsmath}
% \usepackage{graphicx}
% \usepackage[style=apa]{biblatex}
% \usepackage{natbib}
% \usepackage[font=footnotesize,labelfont=bf]{caption}
% \addbibresource{rmt.bib}
% \bibliographystyle{plain}
% \bibliography{rmt}
\begin{document}

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}
\author[Derek M. Berger, Gurpreet S. Matharoo, Jacob Levman]{
    Derek M. Berger\affil{1},
    Gurpreet S. Matharoo\affil{2}
    \and Jacob Levman\affil{3}
}
\affiliation{1}{Computer Science, St. Francis Xavier University, Antigonish, Canada}
\affiliation{2}{Physics, St. Francis Xavier University, Antigonish, Canada}
\affiliation{3}{Computer Science, St. Francis Xavier University, Antigonish, Canada}

% \date{\today}

\keywords{random matrix theory, feature extraction, functional connectivity,
functional magnetic resonance imaging, voxelwise connectivity, predictive analysis}

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse
    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.
\end{abstract}

@@

Dataset scan parameters and acquisition details are summarized in Table \ref{table:1}, and sample
and subgroup sizes are summarized in Table \ref{table:2}. For all datasets, we use the raw
(unpreprocessed) scans, and perform our own preprocessing (detailed later). Given the complexity of
these datasets, we highlight below only those aspects relevant to the current paper.

\begin{table}[h!]
\small
\centering
\begin{tabular}{ c c c c c c c }
\hline
\textbf{ID}  & \textbf{FOV (mm)}  & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{Time (s)} \\
\hline
PSYCH   &  192              & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0 & 300 & 15 \\
OSTEO   &  -                & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5 & 300 & 12.5 \\
REFLECT &  210              & 72 x 72 x 46   & 3.0 x 3.0 x 3.0 & 2.5 & 204 & 10.2 \\
REFLECT &  240              & 64 x 64 x 33   & 3.8 x 3.8 x 3.8 & 2.0 & 260 & 8.7 \\
PARK    &  240 x 240 x 129  & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4 & 149 & 5.96 \\
LEARN   &  -                & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0 & 196 & 6.5 \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:1}
\end{table}

@@

\subsubsection{Dataset 2}
Dataset 2 (referred to in figures and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in figures and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
\citep[ANT;][]{fanActivationAttentionalNetworks2005}. Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

@@

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table \ref{table:1}. The
various RMT features mentioned above were then extracted via the \texttt{empyricalRMT} library
\citep{bergerStfxecutablesEmpyricalRMTPreliminary2020} developed and made publicly available as part
of this study, as well as via code for this paper \citep{bergerderekm.DMBergerRandommatrixfmriV02020}.

@@

\bibliography{rmt}

\section{Technical Terms}

\textbf{Functional Connectivity} the correlations and/or interactions between the timeseries of
various regions of interest in fMRI scans.

\textbf{Random Matrix} A random matrix is a matrix where each entry is sampled from some probability
distribution.

\textbf{Random Matrix Theory} or RMT is an area of mathematics concerned with the properties of
large random matrices with identically and independently distributed (\emph{i.i.d.}) entries. RMT often investigates
the limiting distributions of the eigenvalues of such matrices.

\textbf{Voxelwise Functional Connectivity} the full matrix of correlations and/or interactions
between the timeseries of \emph{all} voxels in an fMRI scan.

\textbf{unfolding} A procedure wherein the eigenvalues of an empirically-observed matrix are
re-mapped to a common space to allow comparison with the spectra of certain known ensembles.

\textbf{Gaussian Orthogonal Ensemble} or GOE. The set of real symmetric matrices with entries
distributed according to the standard Gaussian.

\textbf{The Nearest-Neighbors Spacings Distribution} or NNSD is the is the distribution of the
spacings (differences) of the adjacent unfolded and sorted eigenvalues.

\textbf{Noise Ratio} The proportion of eigenvalues that fall within the maximum and minimum expected
limits of the Marchenko-Pastur distribution.

>>>>

\documentclass[10pt]{article}
\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[english]{babel} % hyphenation
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath}        % https://tex.stackexchange.com/questions/41035/what-is-causing-undefined-control-sequence
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{bm}             % proper bolding
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{verbatim}
% \usepackage{pdflscape}
\usepackage{graphicx}
\usepackage{doi}
\usepackage{cleveref}

\usepackage[round]{natbib}
\setcitestyle{citesep={;}, aysep={,}}
% \bibliographystyle{plainnat}
% NOTE: Must be Better Bibtex for agsm style to work!
\bibliographystyle{agsm}

\newcommand{\tsub}[2]{\mathbf{#1}_{\text{#2}}}
\DeclareMathOperator*{\argmin}{\arg\min}
\graphicspath{ {figs/} }

\begin{document}
\include{meta}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse
    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.
\end{abstract}

\keywords{random matrix theory, feature extraction, functional connectivity,
functional magnetic resonance imaging, voxelwise connectivity, predictive analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Dataset scan parameters and acquisition details are summarized in Table \ref{tab:1}, and sample
and subgroup sizes are summarized in Table \ref{table:2}. For all datasets, we use the raw
(unpreprocessed) scans, and perform our own preprocessing (detailed later). Given the complexity of
these datasets, we highlight below only those aspects relevant to the current paper.

\begin{table}[h!]
  \centering
  \begin{tabular}{ c c c c c c c }
  \hline
  \textbf{ID}  & \textbf{FOV (mm)}  & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{Time (s)} \\
  \hline
  PSYCH   &  192              & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0 & 300 & 15 \\
  OSTEO   &  -                & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5 & 300 & 12.5 \\
  REFLECT &  210              & 72 x 72 x 46   & 3.0 x 3.0 x 3.0 & 2.5 & 204 & 10.2 \\
  REFLECT &  240              & 64 x 64 x 33   & 3.8 x 3.8 x 3.8 & 2.0 & 260 & 8.7 \\
  PARK    &  240 x 240 x 129  & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4 & 149 & 5.96 \\
  LEARN   &  -                & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0 & 196 & 6.5 \\
  \hline
  \end{tabular}
  \caption{
      ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
      Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
      indicate P slices with dimensions M x N.
  }
  \label{tab:1}
\end{table}

@@

\subsubsection{Dataset 2}
Dataset 2 (referred to in figures and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{tab:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in figures and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{tab:1}. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
\citep[ANT;][]{fanActivationAttentionalNetworks2005}. Subgroup details are available in Table \ref{tab:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

@@

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table \ref{tab:1}. The
various RMT features mentioned above were then extracted via the \texttt{empyricalRMT} library
\citep{bergerStfxecutablesEmpyricalRMTPreliminary2020} developed and made publicly available as part
of this study, as well as via code for this paper \citep{bergerderekm.DMBergerRandommatrixfmriV02020}.

@@

\begin{comment}
\hyperref[tab:existing-attempts]{Table \ref{tab:existing-attempts}}).
\begin{table}
	\centering
	\begin{tabular}{lll}
		\toprule
		Feature     & Name     & Shape  \\
		\midrule
    ROI means                                                                                  & \(\tsub{\mu}{ROI}\)                        & \((n, T)\)    \\
    ROI standard deviations (sds)                                                              & \(\tsub{\sigma}{ROI}\)                     & \((n, T)\)    \\
    ROI descriptive sequences                                                                  & \((\tsub{\mu}{ROI}, \tsub{\sigma}{ROI})\)  & \((2, n, T)\) or \((2n, T)\) \\
    Correlation of ROI means                                                                   & \(\mathbf{r}_{\mu,\text{ROI}}\)            & \((n, n)\)    \\
    Correlation of ROI sds                                                                     & \(\mathbf{r}_{\sigma,\text{ROI}}\)         & \((n, n)\)    \\
    Correlation descriptives \([\mathbf{r}_{\mu,\text{ROI}}; \mathbf{r}_{\sigma\text{,ROI}}]\) & \(\tsub{r}{ROI}\)                          & \((2, n, n)\) or \((2n, n)\)\\
    T-thresholded Laplacian eigenvalues of \(\mathbf{r}_{\mu,\text{ROI}}\)                     & \(\mathbf{\lambda}_{\mu,\text{ROI},T}\)    & \(n-1\)       \\
    T-thresholded Laplacian eigenvalues of \(\mathbf{r}_{\sigma\text{,ROI}}\)                  & \(\mathbf{\lambda}_{\sigma,\text{ROI},T}\) & \(n-1\)       \\
    Concatenated T-thresholded Laplacian eigenvalues                                           & \(\mathbf{\lambda}_{\text{ROI},T}\)        & \((2, n-1)\)  \\
    Eigenvalues of \(\mathbf{r}_{\mu,\text{ROI}}\)                                             & \(\mathbf{\lambda}_{\mu,\text{ROI}}\)      & \(n-1\)       \\
    Eigenvalues of \(\mathbf{r}_{\sigma\text{,ROI}}\)                                          & \(\mathbf{\lambda}_{\sigma,\text{ROI}}\)   & \(n-1\)       \\
    Concatenated eigenvalues \([\mathbf{r}_{\sigma\text{,ROI}}; \mathbf{r}_{\sigma\text{,ROI}}]\)  & \(\tsub{\lambda}{ROI}\)                & \((2, n-1)\)  \\
		\bottomrule
	\end{tabular}
	\label{tab:feature-shapes}
	\caption{ROI-based feature shapes. \(n\) = number of ROIs. \([A;B]\) = concatenation of \(A\) and \(B\) along new initial (channel) dimension for deep learning, or along initial dimension, for classical ML.}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{rr}
  \toprule
  T &   \(N\) subjects \\
  \midrule
   78 &   26 \\
  116 &  134 \\
  124 &    6 \\
  146 &   84 \\
  152 &   42 \\
  176 &  227 \\
  177 &    1 \\
  196 &  158 \\
  202 &    1 \\
  206 &   34 \\
  232 &    1 \\
  234 &    2 \\
  236 &  103 \\
  246 &   63 \\
  296 &  140 \\
  316 &   13 \\
  \bottomrule
\end{tabular}\label{tab:timepoints}
\caption{Heterogeneity in ABIDE number of timepoints. Mean number of timepoints = 193, median = 176. Standard deviation = 58.1}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{cr}
  \toprule
  TR (s) &   \(N\) subjects \\
  \midrule
  1.5  &      81 \\
  1.7 &       56 \\
  2    &     519 \\
  2.2  &      26 \\
  2.5  &      58 \\
  3    &     131 \\
  \bottomrule
  \end{tabular}\label{tab:trs}
\caption{Heterogeneity in ABIDE repetition times. Note many subjects have different TR values very close to 1.666 due to rounding issue, so are just reported here as 1.7 for simplicity}
\end{table}
\end{comment}

% \bibliographystyle{unsrtnat}
\bibliography{notes}
